---
title: "RStataLink -- R package for calling Stata from R interactively"
author: "Aleksander Rutkowski"
date: '`r format(Sys.Date(), "%d %B %Y")`'
output:
  html_document:
    keep_md: true
---

## Features

Execute smaller or larger bits and pieces of Stata code **interactively** from R 
in a [Stata](http://www.stata.com/) "server" i.e. not in a batch mode:

- avoid repeated costly operations, e.g. open Stata and load data in Stata **only once**
and do subsequent operations requiring R<->Stata interaction without
re-opening and re-loading
- easy R<->Stata **data exchange** (import/export)
- easy Stata r() and e() **"results" extraction** (macros, scalars, matrices, coefficients
and standard errors), shaped for convenient consumption in R
- **concurrency** "primitive" ([future](https://en.wikipedia.org/wiki/Futures_and_promises)-like)
and parallel `apply()`-style functions for executing Stata code in a concurrent/parallel
manner in a "cluster" of multiple Stata instances (regular and load-balancing versions)
- should work **cross-platform** -- pure R and pure Stata code (but developed and tested on
MS Windows only)

Make your work with Stata more
[functional](https://en.wikipedia.org/wiki/Functional_programming)!

Similar but unrelated project: <https://github.com/lbraglia/RStata>. It seems
to offer only batch mode i.e. not interactive.

## How it works

Stata "server" is a Stata instance running an infinite loop and waiting for new jobs
showing up in a specific directory/folder. Thus, the R<->Stata communication is
disk-based so it can take place only locally (within a single computer) or through a
shared network drive (across computers).

In the latter case (the shared drive approach), Stata would still need to be
opened by R with `startStata()` or `startStataCluster()` on the "server" computer
(e.g. via SSH). The generated `StataID` object(s) (see `i` or `cl`
below) need(s) to be serialised (e.g. with `saveRDS()`) and transmitted in some way
to a remote "client" computer to be deserialised (e.g. with `readRDS()`).
The entire path to the shared network drive directory (see argument `compath` in
`startStata()`) should be the same on both computers ("server" and "client").

## Installation

```{r, eval=FALSE, include=TRUE}
if ('devtools' %in% installed.packages()[,"Package"])
	devtools::install_github('alekrutkowski/RStataLink') else
	stop('You need package "devtools"!')
```

## Usage examples

Tell R where Stata is (can be also per-call -- using argument `start_cmd` in `startStata()` --
but that would be less convenient):

```{r}
# A virtualised app in my case, therefore such a complicated path, should be simpler normally:
options(statapath=paste('"C:\\Program Files (x86)\\Microsoft Application Virtualization Client\\sfttray.exe"',
                        '/launch "StataMP 14 (64-bit) [V] 14.0bd001"'))
```

### A single Stata instance functionality demo

For the Stata code pieces use a string with a newline
character (\\n), or multi-line string, or a character vector (that will be converted to a
multi-line string):

```{r, comment="##     "}
library(RStataLink)
i <- startStata()
i
r1 <- doInStata(i, 'display 100 + 15.7', results = NULL)
r1  # results = NULL means no import of Stata e() or r() results
# Use in Stata a built-in demo dataset on cars
# and make a simple regression with robust standard errors
# (the 3 lines of code could be also done with 3 consecutive doInStata calls)
r2 <- doInStata(i,
				'sysuse auto, clear
				regress price weight trunk, robust
				ereturn list')
r2$log
r2$results  # this (below) looks similar to "ereturn list" in Stata (above)
# Use in Stata an R demo dataset on flowers
# modify it (temporarily in Stata with preserve...restore,
# see http://www.stata.com/help.cgi?preserve)
# and do a simple regression:
data(iris)
r3 <- doInStata(i, code = 
                'describe
				gen ln_sepallength = log(sepallength)
				reg ln_sepallength sepalwidth petallength petalwidth',
				df = iris,
				preserve_restore = TRUE)
r3$log
# If data is exported to Stata with df = ...
# a (possibly modified) Stata dataset is imported back
# into R unless argument import_df = FALSE is
# specified in doInStata()
str(r3$df)
r3$results$e_class  # again, the estimated results from e() are available:
# Since we did preserve...restore in Stata
# while operating on the iris data,
# the data on cars is still there
doInStata(i, 'describe', results=NULL)
# Also r-class Stata results can be collected:
r4 <- doInStata(i, 'summarize price \n return list',
				results = 'r')  # Stata r-class results only
r4
# A non-blocking call to Stata -- perform some long-running job
system.time({f <- doInStata(i, 'sleep 8000 // in milliseconds in Stata
							display "hello"',
							future = TRUE,
							results = NULL)})
# do some work in R in the meantime:
Sys.sleep(2)
# collect the results from Stata if ready
# (if not ready, you must wait -- it's blocking this time,
# note the approximate time difference: 8 - 2 = 6):
system.time({r5 <- getStataFuture(f)})
r5
# Say good-bye to Stata:
stopStata(i)
```

### A multiple Stata instance functionality demo

```{r, comment="##     "}
library(RStataLink)
# The length of cl (the number of Stata instances) will depend
# on the number of cores detected by parallel::detectCores()
# (but you can override it)
cl <- startStataCluster()
# It's just a simple list overall:
class(cl)
# Have a look at the first element:
cl[[1]]
# A trivial example - a series of different regressions with
# only one explanatory variable each:
m <- paste('regress sepallength',
				 c('sepalwidth',
				   'petallength',
				   'petalwidth'))
cat(m, sep = '\n')
c1 <- doInStataCluster(cl,
					   # X = a vector (atomic or list) of tasks/jobs
					   # expressed as a character vector
					   # or a list of character vectors (which will be
					   # collapsed into string each)
					   X = m,
					   nolog = TRUE,
					   df = iris,
					   import_df = FALSE)
# A load-balancing version -- use if some jobs/tasks
# are much longer than others (iris data already loaded
# so no need for df = iris):
c2 <- doInStataClusterLB(cl,
					   X = m,
					   nolog = TRUE)
identical(c1, c2)
# Collect the results:
library(magrittr) # for the super cool pipe operator %>%
c1 %>%
	lapply(function(x) x$results$e_class$modeldf) %>%
	do.call(rbind, .)
# Say good-bye to all the Statas opened by R
# (clear = TRUE so that each Stata allows you
# to discard the imported iris data):
stopStataCluster(cl, clear = TRUE)
```

